---
title: "NFL Player Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages

# install.packages("nflreadr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr")

library(nflreadr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which player do you want to grade?
# Use the player's full name as NFL rosters list it (First Last).
# Examples: "Patrick Mahomes", "Josh Allen", "Tyreek Hill", "Derrick Henry"
my_player <- "Patrick Mahomes"

# Which season? NFL uses the year the season STARTS.
# Example: the 2024-25 season is written as 2024.
season_year <- 2024

# Season label shown in the report (for display only)
season_label <- "2024-25"

# Player position — determines which stats and benchmarks are used.
# Choose ONE of:  "QB"  |  "RB"  |  "WR"  |  "TE"
player_position <- "QB"

# Include playoffs? TRUE = regular season + playoffs, FALSE = regular season only
include_playoffs <- FALSE

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

stat_type <- if (player_position == "QB") "passing" else "receiving"
if (player_position %in% c("RB")) stat_type <- "rushing"

# Load both offense types and merge
pass_stats <- nflreadr::load_player_stats(seasons = season_year, stat_type = "offense")

if (!include_playoffs) {
  pass_stats <- pass_stats %>% filter(season_type == "REG")
}

# Find the player
player_rows <- pass_stats %>%
  filter(tolower(player_display_name) == tolower(my_player))

if (nrow(player_rows) == 0) {
  player_rows <- pass_stats %>%
    filter(grepl(my_player, player_display_name, ignore.case = TRUE))
}

if (nrow(player_rows) == 0) {
  stop("Player '", my_player, "' not found. Run: ",
       "nflreadr::load_player_stats(seasons=", season_year, ", stat_type='offense') %>% ",
       "filter(grepl('LAST_NAME', player_display_name, ignore.case=TRUE)) %>% ",
       "distinct(player_display_name) to find the exact name.")
}

player_name_full <- player_rows$player_display_name[1]
team_name_full   <- player_rows$recent_team[1]
games_played     <- nrow(player_rows)
```

```{r}
#| label: averages
#| include: false

avg <- player_rows %>%
  summarise(
    # Passing (QB)
    completions     = mean(completions,      na.rm = TRUE),
    attempts        = mean(attempts,          na.rm = TRUE),
    pass_yards      = mean(passing_yards,     na.rm = TRUE),
    pass_tds        = mean(passing_tds,       na.rm = TRUE),
    interceptions   = mean(interceptions,     na.rm = TRUE),
    passer_rating   = mean(passer_rating,     na.rm = TRUE),
    # Rushing
    carries         = mean(carries,           na.rm = TRUE),
    rush_yards      = mean(rushing_yards,     na.rm = TRUE),
    rush_tds        = mean(rushing_tds,       na.rm = TRUE),
    # Receiving
    receptions      = mean(receptions,        na.rm = TRUE),
    targets         = mean(targets,           na.rm = TRUE),
    recv_yards      = mean(receiving_yards,   na.rm = TRUE),
    recv_tds        = mean(receiving_tds,     na.rm = TRUE),
    # Shared
    fumbles_lost    = mean(sack_fumbles_lost + rushing_fumbles_lost +
                             receiving_fumbles_lost, na.rm = TRUE),
    fantasy_points  = mean(fantasy_points,    na.rm = TRUE)
  ) %>%
  mutate(
    comp_pct   = completions / pmax(attempts, 1) * 100,
    catch_rate = receptions  / pmax(targets,  1) * 100
  )

# Season totals
totals <- player_rows %>%
  summarise(
    total_pass_yards = sum(passing_yards,   na.rm = TRUE),
    total_pass_tds   = sum(passing_tds,     na.rm = TRUE),
    total_int        = sum(interceptions,   na.rm = TRUE),
    total_rush_yards = sum(rushing_yards,   na.rm = TRUE),
    total_rush_tds   = sum(rushing_tds,     na.rm = TRUE),
    total_recv_yards = sum(receiving_yards, na.rm = TRUE),
    total_recv_tds   = sum(receiving_tds,   na.rm = TRUE),
    total_rec        = sum(receptions,      na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (is.na(value) || is.nan(value)) return("N/A")
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  if (l == "N/A") return(NA_real_)
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

if (player_position == "QB") {

  grade_pass_yards  <- grade_stat(avg$pass_yards,    c(A=275, B=235, C=200, D=165))
  grade_pass_tds    <- grade_stat(avg$pass_tds,      c(A=2.2, B=1.7, C=1.2, D=0.8))
  grade_comp_pct    <- grade_stat(avg$comp_pct,      c(A=68,  B=64,  C=60,  D=56))
  grade_ints        <- grade_stat(avg$interceptions, c(A=0.5, B=0.8, C=1.2, D=1.7), lower_is_better=TRUE)
  grade_qbr         <- grade_stat(avg$passer_rating, c(A=105, B=95,  C=85,  D=75))
  grade_mobility    <- grade_stat(avg$rush_yards,    c(A=30,  B=20,  C=12,  D=5))

  all_grades <- c(grade_pass_yards, grade_pass_tds, grade_comp_pct,
                  grade_ints, grade_qbr, grade_mobility)
  weights    <- c(0.22, 0.22, 0.18, 0.20, 0.12, 0.06)

  stat_categories <- c("Pass Yards/G","Pass TDs/G","Completion %","INTs/G","Passer Rating","Rush Yards/G")
  stat_values     <- c(sprintf("%.1f", avg$pass_yards),
                       sprintf("%.2f", avg$pass_tds),
                       sprintf("%.1f%%", avg$comp_pct),
                       sprintf("%.2f", avg$interceptions),
                       sprintf("%.1f", avg$passer_rating),
                       sprintf("%.1f", avg$rush_yards))

} else if (player_position == "RB") {

  grade_rush_yards  <- grade_stat(avg$rush_yards,    c(A=85, B=65, C=48, D=30))
  grade_rush_tds    <- grade_stat(avg$rush_tds,      c(A=0.8, B=0.5, C=0.3, D=0.15))
  grade_receptions  <- grade_stat(avg$receptions,    c(A=5, B=3.5, C=2.2, D=1.2))
  grade_recv_yards  <- grade_stat(avg$recv_yards,    c(A=45, B=30, C=18, D=8))
  grade_fumbles     <- grade_stat(avg$fumbles_lost,  c(A=0.05, B=0.12, C=0.20, D=0.30), lower_is_better=TRUE)

  all_grades <- c(grade_rush_yards, grade_rush_tds, grade_receptions,
                  grade_recv_yards, grade_fumbles)
  weights    <- c(0.30, 0.25, 0.18, 0.17, 0.10)

  stat_categories <- c("Rush Yards/G","Rush TDs/G","Receptions/G","Recv Yards/G","Fumbles Lost/G")
  stat_values     <- c(sprintf("%.1f", avg$rush_yards),
                       sprintf("%.2f", avg$rush_tds),
                       sprintf("%.1f", avg$receptions),
                       sprintf("%.1f", avg$recv_yards),
                       sprintf("%.2f", avg$fumbles_lost))

} else {  # WR or TE

  grade_recv_yards  <- grade_stat(avg$recv_yards,   c(A=75, B=55, C=38, D=22))
  grade_recv_tds    <- grade_stat(avg$recv_tds,      c(A=0.6, B=0.4, C=0.25, D=0.12))
  grade_receptions  <- grade_stat(avg$receptions,   c(A=6.5, B=5,   C=3.5,  D=2))
  grade_catch_rate  <- grade_stat(avg$catch_rate,   c(A=72,  B=65,  C=58,   D=50))
  grade_ypr         <- grade_stat(avg$recv_yards / pmax(avg$receptions, 1),
                                   c(A=15, B=12, C=9, D=7))

  all_grades <- c(grade_recv_yards, grade_recv_tds, grade_receptions,
                  grade_catch_rate, grade_ypr)
  weights    <- c(0.28, 0.25, 0.20, 0.15, 0.12)

  stat_categories <- c("Recv Yards/G","Recv TDs/G","Receptions/G","Catch Rate","Yds/Reception")
  stat_values     <- c(sprintf("%.1f", avg$recv_yards),
                       sprintf("%.2f", avg$recv_tds),
                       sprintf("%.1f", avg$receptions),
                       sprintf("%.1f%%", avg$catch_rate),
                       sprintf("%.1f", avg$recv_yards / pmax(avg$receptions, 1)))
}

valid      <- !is.na(sapply(all_grades, letter_to_gpa))
overall_gpa   <- weighted.mean(sapply(all_grades, letter_to_gpa)[valid], weights[valid])
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = stat_categories,
  Value    = stat_values,
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
) %>% filter(!is.na(GPA))

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r player_name_full` — `r team_name_full` | `r season_label`

**Position:** `r player_position` | **Games Played:** `r games_played`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(player_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NFL Season |", player_position),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Week-by-Week Performance

```{r}
#| label: weekly-chart

weekly_data <- player_rows %>%
  arrange(week) %>%
  mutate(
    primary_yards = case_when(
      player_position == "QB" ~ passing_yards,
      player_position == "RB" ~ rushing_yards,
      TRUE                    ~ receiving_yards
    ),
    y_label = case_when(
      player_position == "QB" ~ "Passing Yards",
      player_position == "RB" ~ "Rushing Yards",
      TRUE                    ~ "Receiving Yards"
    )
  )

y_label <- case_when(
  player_position == "QB" ~ "Passing Yards",
  player_position == "RB" ~ "Rushing Yards",
  TRUE                    ~ "Receiving Yards"
)

ggplot(weekly_data, aes(x = week, y = primary_yards)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_hline(yintercept = mean(weekly_data$primary_yards, na.rm=TRUE),
             linetype="dashed", color="#e74c3c", linewidth=0.8) +
  labs(
    title    = paste(player_name_full, "—", y_label, "by Week"),
    subtitle = "Red dashed line = season average",
    x        = "Week",
    y        = y_label
  ) +
  theme_minimal(base_size = 13)
```

## Season Totals Summary

```{r}
#| label: stats-table

if (player_position == "QB") {
  summary_tbl <- data.frame(
    Statistic = c("Games Played","Passing Yards","Passing TDs","Interceptions",
                  "Completion %","Passer Rating","Rush Yards"),
    Value     = c(games_played,
                  totals$total_pass_yards,
                  totals$total_pass_tds,
                  totals$total_int,
                  sprintf("%.1f%%", avg$comp_pct),
                  sprintf("%.1f", avg$passer_rating),
                  totals$total_rush_yards)
  )
} else if (player_position == "RB") {
  summary_tbl <- data.frame(
    Statistic = c("Games Played","Rushing Yards","Rushing TDs",
                  "Receptions","Receiving Yards","Receiving TDs"),
    Value     = c(games_played, totals$total_rush_yards, totals$total_rush_tds,
                  totals$total_rec, totals$total_recv_yards, totals$total_recv_tds)
  )
} else {
  summary_tbl <- data.frame(
    Statistic = c("Games Played","Receptions","Receiving Yards","Receiving TDs",
                  "Catch Rate","Yards Per Reception"),
    Value     = c(games_played, totals$total_rec, totals$total_recv_yards,
                  totals$total_recv_tds,
                  sprintf("%.1f%%", avg$catch_rate),
                  sprintf("%.1f", totals$total_recv_yards / pmax(totals$total_rec, 1)))
  )
}

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season"))
```

## How the Grade Is Calculated

Grades are calibrated to the player's position — a QB is not graded on the same yardage scale as a wide receiver. Each category is compared against typical NFL starter averages for that position and converted to a 4.0 scale. For QBs, interceptions and passer rating carry heavy weight because they reflect efficiency and decision-making. For skill players, yardage totals and touchdowns are the primary drivers.
