---
title: "MLB Player Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages

# install.packages("baseballr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr")

library(baseballr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which player do you want to grade?
# Use the player's full name as FanGraphs lists it (First Last).
# Examples: "Aaron Judge", "Shohei Ohtani", "Mookie Betts", "Gerrit Cole"
my_player <- "Aaron Judge"

# Which season? Use the calendar year.
season_year <- 2024

# Season label shown in the report (for display only)
season_label <- "2024"

# Player type — determines which stats and benchmarks are used:
# "Batter"  for position players
# "Pitcher" for starting and relief pitchers
player_type <- "Batter"

# Minimum plate appearances / innings to qualify (used to filter leaderboard)
# Batter default: 300 PA  |  Pitcher default: 50 IP
min_qualifier <- 300

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

if (player_type == "Batter") {

  leaders <- baseballr::fg_batter_leaders(
    startseason = as.character(season_year),
    endseason   = as.character(season_year),
    qual        = min_qualifier
  )

  player_data <- leaders %>%
    filter(tolower(playerName) == tolower(my_player))

  if (nrow(player_data) == 0) {
    # Try partial match / lower qualifier
    all_batters <- baseballr::fg_batter_leaders(
      startseason = as.character(season_year),
      endseason   = as.character(season_year),
      qual        = 1
    )
    player_data <- all_batters %>%
      filter(grepl(my_player, playerName, ignore.case = TRUE))
  }

  if (nrow(player_data) == 0) {
    stop("Batter '", my_player, "' not found. Run: ",
         "baseballr::fg_batter_leaders(startseason='", season_year,
         "', endseason='", season_year, "', qual=1) %>% ",
         "filter(grepl('LAST_NAME', playerName, ignore.case=TRUE)) to find the name.")
  }

  player_data    <- player_data %>% slice(1)
  player_name_full <- player_data$playerName[1]
  team_name_full   <- player_data$team[1]

} else {

  leaders <- baseballr::fg_pitcher_leaders(
    startseason = as.character(season_year),
    endseason   = as.character(season_year),
    qual        = min_qualifier
  )

  player_data <- leaders %>%
    filter(tolower(playerName) == tolower(my_player))

  if (nrow(player_data) == 0) {
    all_pitchers <- baseballr::fg_pitcher_leaders(
      startseason = as.character(season_year),
      endseason   = as.character(season_year),
      qual        = 1
    )
    player_data <- all_pitchers %>%
      filter(grepl(my_player, playerName, ignore.case = TRUE))
  }

  if (nrow(player_data) == 0) {
    stop("Pitcher '", my_player, "' not found.")
  }

  player_data    <- player_data %>% slice(1)
  player_name_full <- player_data$playerName[1]
  team_name_full   <- player_data$team[1]
}
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (is.na(value) || is.nan(value)) return("N/A")
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  if (l == "N/A") return(NA_real_)
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

if (player_type == "Batter") {

  avg_val  <- as.numeric(player_data$AVG)
  obp      <- as.numeric(player_data$OBP)
  slg      <- as.numeric(player_data$SLG)
  ops      <- as.numeric(player_data$OPS)
  woba     <- as.numeric(player_data$wOBA)
  wrc_plus <- as.numeric(player_data$wRC_plus)   # column may be wRC+ or wRCplus
  hr       <- as.numeric(player_data$HR)
  rbi      <- as.numeric(player_data$RBI)
  sb       <- as.numeric(player_data$SB)
  bb_pct   <- as.numeric(player_data$`BB%`) * 100
  k_pct    <- as.numeric(player_data$`K%`)  * 100
  games    <- as.numeric(player_data$G)
  pa       <- as.numeric(player_data$PA)

  # Benchmarks: MLB qualified batter averages (2024 season)
  grade_avg    <- grade_stat(avg_val,  c(A=.290, B=.270, C=.250, D=.230))
  grade_obp    <- grade_stat(obp,      c(A=.360, B=.335, C=.310, D=.290))
  grade_slg    <- grade_stat(slg,      c(A=.500, B=.450, C=.400, D=.360))
  grade_ops    <- grade_stat(ops,      c(A=.860, B=.790, C=.720, D=.650))
  grade_woba   <- grade_stat(woba,     c(A=.380, B=.345, C=.310, D=.275))
  grade_hr     <- grade_stat(hr,       c(A=35,   B=25,   C=15,   D=8))
  grade_bb     <- grade_stat(bb_pct,   c(A=12,   B=9,    C=7,    D=5))
  grade_k      <- grade_stat(k_pct,    c(A=14,   B=18,   C=23,   D=28),  lower_is_better=TRUE)

  all_grades <- c(grade_avg, grade_obp, grade_slg, grade_ops,
                  grade_woba, grade_hr, grade_bb, grade_k)
  weights    <- c(0.10, 0.14, 0.12, 0.18, 0.18, 0.12, 0.08, 0.08)

  stat_categories <- c("Batting Avg","OBP","SLG","OPS","wOBA","Home Runs","Walk %","Strikeout %")
  stat_values     <- c(
    sprintf(".%03d", round(avg_val * 1000)),
    sprintf(".%03d", round(obp     * 1000)),
    sprintf(".%03d", round(slg     * 1000)),
    sprintf(".%03d", round(ops     * 1000)),
    sprintf(".%03d", round(woba    * 1000)),
    as.character(as.integer(hr)),
    sprintf("%.1f%%", bb_pct),
    sprintf("%.1f%%", k_pct)
  )

} else {

  era    <- as.numeric(player_data$ERA)
  whip   <- as.numeric(player_data$WHIP)
  fip    <- as.numeric(player_data$FIP)
  xfip   <- as.numeric(player_data$xFIP)
  k9     <- as.numeric(player_data$`K/9`)
  bb9    <- as.numeric(player_data$`BB/9`)
  k_pct  <- as.numeric(player_data$`K%`) * 100
  ip     <- as.numeric(player_data$IP)
  wins   <- as.numeric(player_data$W)
  sv     <- as.numeric(player_data$SV)
  games  <- as.numeric(player_data$G)

  # Benchmarks: MLB qualified starter averages (2024 season)
  grade_era  <- grade_stat(era,   c(A=2.8, B=3.5, C=4.2, D=5.0), lower_is_better=TRUE)
  grade_whip <- grade_stat(whip,  c(A=1.05,B=1.18,C=1.30,D=1.42), lower_is_better=TRUE)
  grade_fip  <- grade_stat(fip,   c(A=2.9, B=3.5, C=4.1, D=4.8), lower_is_better=TRUE)
  grade_k9   <- grade_stat(k9,    c(A=11,  B=9.5, C=8,   D=6.5))
  grade_bb9  <- grade_stat(bb9,   c(A=1.8, B=2.4, C=3.0, D=3.8), lower_is_better=TRUE)
  grade_kpct <- grade_stat(k_pct, c(A=30,  B=25,  C=20,  D=15))

  all_grades <- c(grade_era, grade_whip, grade_fip, grade_k9, grade_bb9, grade_kpct)
  weights    <- c(0.22, 0.18, 0.20, 0.16, 0.12, 0.12)

  stat_categories <- c("ERA","WHIP","FIP","K/9","BB/9","K%")
  stat_values     <- c(sprintf("%.2f", era), sprintf("%.2f", whip),
                       sprintf("%.2f", fip), sprintf("%.1f", k9),
                       sprintf("%.1f", bb9), sprintf("%.1f%%", k_pct))
}

valid      <- !is.na(sapply(all_grades, letter_to_gpa))
overall_gpa   <- weighted.mean(sapply(all_grades, letter_to_gpa)[valid], weights[valid])
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = stat_categories,
  Value    = stat_values,
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
) %>% filter(!is.na(GPA))

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r player_name_full` — `r team_name_full` | `r season_label`

**Type:** `r player_type` | **Games:** `r games` `r if(player_type=="Batter") paste0("| PA: ", pa) else paste0("| IP: ", ip)`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(player_name_full, "— Category Grades"),
    subtitle = paste(season_label, "MLB Season |", player_type),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Stat Comparison vs. MLB Average

```{r}
#| label: comparison-chart

if (player_type == "Batter") {
  compare_df <- data.frame(
    Stat     = c("AVG","OBP","SLG","OPS","wOBA"),
    Player   = c(avg_val, obp, slg, ops, woba),
    MLB_Avg  = c(.243, .310, .394, .704, .313)
  )
} else {
  compare_df <- data.frame(
    Stat     = c("ERA","WHIP","FIP","K/9"),
    Player   = c(era, whip, fip, k9),
    MLB_Avg  = c(4.20, 1.28, 4.10, 8.8)
  )
}

compare_long <- compare_df %>%
  pivot_longer(cols = c(Player, MLB_Avg), names_to = "Source", values_to = "Value") %>%
  mutate(Source = recode(Source, Player = player_name_full, MLB_Avg = "MLB Average"))

ggplot(compare_long, aes(x = Stat, y = Value, fill = Source)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("#3498db","#bdc3c7")) +
  geom_text(aes(label = sprintf("%.3f", Value)),
            position = position_dodge(width=0.6), vjust=-0.4, size=3.2) +
  labs(
    title    = paste(player_name_full, "vs. MLB Average"),
    subtitle = season_label,
    x        = NULL, y = "Value", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Season Stats Summary

```{r}
#| label: stats-table

if (player_type == "Batter") {
  summary_tbl <- data.frame(
    Statistic = c("Games","Plate Appearances","Batting Average","OBP",
                  "SLG","OPS","wOBA","Home Runs","RBI","Stolen Bases",
                  "Walk %","Strikeout %"),
    Value     = c(
      as.integer(games), as.integer(pa),
      sprintf(".%03d", round(avg_val * 1000)),
      sprintf(".%03d", round(obp     * 1000)),
      sprintf(".%03d", round(slg     * 1000)),
      sprintf(".%03d", round(ops     * 1000)),
      sprintf(".%03d", round(woba    * 1000)),
      as.integer(hr), as.integer(rbi), as.integer(sb),
      sprintf("%.1f%%", bb_pct), sprintf("%.1f%%", k_pct)
    )
  )
} else {
  summary_tbl <- data.frame(
    Statistic = c("Games","Innings Pitched","ERA","WHIP","FIP","xFIP",
                  "K/9","BB/9","K%","Wins","Saves"),
    Value     = c(
      as.integer(games), ip,
      sprintf("%.2f", era), sprintf("%.2f", whip),
      sprintf("%.2f", fip), sprintf("%.2f", xfip),
      sprintf("%.1f", k9), sprintf("%.1f", bb9),
      sprintf("%.1f%%", k_pct),
      as.integer(wins), as.integer(sv)
    )
  )
}

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season"))
```

## How the Grade Is Calculated

For batters, OPS and wOBA each carry 18% of the overall grade because they are the most complete measures of offensive value — OPS is familiar to fans, while wOBA weights each type of hit by its actual run value. For pitchers, ERA and FIP are weighted most heavily. FIP (Fielding Independent Pitching) isolates what the pitcher actually controlled — strikeouts, walks, and home runs — removing the influence of the defense behind them.

Data comes from FanGraphs via the `baseballr` package.
