---
title: "NFL Team Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages

# install.packages("nflreadr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr")

library(nflreadr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which NFL team do you want to grade?
# Use the 2-3 letter team abbreviation.
# Examples: "NE", "KC", "DAL", "SF", "BUF", "PHI", "MIA", "LAR"
# Not sure? Run: nflreadr::load_teams() %>% select(team_abbr, team_name) %>% print(n=40)
my_team <- "KC"

# Which season? NFL uses the year the season STARTS.
# Example: the 2024-25 season is written as 2024.
season_year <- 2024

# Season label shown in the report (for display only)
season_label <- "2024-25"

# Include playoffs? TRUE = regular season + playoffs, FALSE = regular season only
include_playoffs <- FALSE

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

schedules <- nflreadr::load_schedules(seasons = season_year)

if (!include_playoffs) {
  schedules <- schedules %>% filter(game_type == "REG")
}

# Home games
home_games <- schedules %>%
  filter(home_team == my_team, !is.na(home_score)) %>%
  transmute(
    game_id,
    week,
    opponent     = away_team,
    team_score   = home_score,
    opp_score    = away_score,
    won          = home_score > away_score,
    home_away    = "Home"
  )

# Away games
away_games <- schedules %>%
  filter(away_team == my_team, !is.na(away_score)) %>%
  transmute(
    game_id,
    week,
    opponent     = home_team,
    team_score   = away_score,
    opp_score    = home_score,
    won          = away_score > home_score,
    home_away    = "Away"
  )

team_data <- bind_rows(home_games, away_games) %>% arrange(week)

if (nrow(team_data) == 0) stop("No games found for '", my_team, "'. Check the abbreviation.")

# Pull offensive and defensive team stats from play-by-play
pbp <- nflreadr::load_pbp(seasons = season_year)

if (!include_playoffs) {
  pbp <- pbp %>% filter(season_type == "REG")
}

# Offensive stats (when team is posteam)
off_stats <- pbp %>%
  filter(posteam == my_team, !is.na(yards_gained)) %>%
  group_by(game_id) %>%
  summarise(
    pass_yards  = sum(passing_yards, na.rm = TRUE),
    rush_yards  = sum(rushing_yards, na.rm = TRUE),
    turnovers   = sum(interception + fumble_lost, na.rm = TRUE),
    sacks_taken = sum(sack, na.rm = TRUE),
    .groups = "drop"
  )

# Defensive stats (when team is defteam)
def_stats <- pbp %>%
  filter(defteam == my_team, !is.na(yards_gained)) %>%
  group_by(game_id) %>%
  summarise(
    sacks_made  = sum(sack, na.rm = TRUE),
    def_tov     = sum(interception + fumble_lost, na.rm = TRUE),
    .groups = "drop"
  )

team_data <- team_data %>%
  left_join(off_stats, by = "game_id") %>%
  left_join(def_stats, by = "game_id")

team_name_full <- nflreadr::load_teams() %>%
  filter(team_abbr == my_team) %>%
  pull(team_name) %>% first()

games_played <- nrow(team_data)
wins         <- sum(team_data$won, na.rm = TRUE)
losses       <- games_played - wins
win_pct      <- wins / games_played * 100
```

```{r}
#| label: averages
#| include: false

avg <- team_data %>%
  summarise(
    pts_for     = mean(team_score,   na.rm = TRUE),
    pts_against = mean(opp_score,    na.rm = TRUE),
    pass_yards  = mean(pass_yards,   na.rm = TRUE),
    rush_yards  = mean(rush_yards,   na.rm = TRUE),
    turnovers   = mean(turnovers,    na.rm = TRUE),
    sacks_taken = mean(sacks_taken,  na.rm = TRUE),
    sacks_made  = mean(sacks_made,   na.rm = TRUE),
    def_tov     = mean(def_tov,      na.rm = TRUE),
    pt_diff     = mean(team_score - opp_score, na.rm = TRUE),
    total_yards = mean(pass_yards + rush_yards, na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (is.na(value)) return("N/A")
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  if (l == "N/A") return(NA_real_)
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# Benchmarks: NFL per-game averages (2024 season)
grade_scoring   <- grade_stat(avg$pts_for,    c(A=28, B=24, C=20, D=17))
grade_defense   <- grade_stat(avg$pts_against, c(A=17, B=20, C=24, D=28), lower_is_better=TRUE)
grade_passing   <- grade_stat(avg$pass_yards, c(A=280, B=245, C=210, D=175))
grade_rushing   <- grade_stat(avg$rush_yards, c(A=145, B=120, C=95,  D=75))
grade_off_tov   <- grade_stat(avg$turnovers,  c(A=0.8, B=1.2, C=1.6, D=2.2), lower_is_better=TRUE)
grade_sacks     <- grade_stat(avg$sacks_made,  c(A=3.5, B=2.7, C=2.0, D=1.3))
grade_def_tov   <- grade_stat(avg$def_tov,    c(A=2.0, B=1.5, C=1.0, D=0.6))
grade_win_pct   <- grade_stat(win_pct,        c(A=70, B=60, C=50, D=38))

all_grades <- c(grade_scoring, grade_defense, grade_passing, grade_rushing,
                grade_off_tov, grade_sacks, grade_def_tov, grade_win_pct)
weights    <- c(0.18, 0.18, 0.12, 0.10, 0.12, 0.10, 0.10, 0.10)

valid      <- !is.na(sapply(all_grades, letter_to_gpa))
overall_gpa   <- weighted.mean(sapply(all_grades, letter_to_gpa)[valid], weights[valid])
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Scoring","Defense (pts allowed)","Passing Yards",
               "Rushing Yards","Turnovers (off)","Sacks (def)",
               "Takeaways (def)","Win %"),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
) %>% filter(!is.na(GPA))

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r team_name_full` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) | **Point Differential:** `r sprintf("%+.1f", avg$pt_diff)` per game | **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(team_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NFL Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Scoring Trend

```{r}
#| label: scoring-trend

trend_data <- team_data %>%
  mutate(result = ifelse(won, "Win", "Loss"))

ggplot(trend_data, aes(x = week)) +
  geom_line(aes(y = team_score), color = "#2ecc71", linewidth = 1) +
  geom_line(aes(y = opp_score),  color = "#e74c3c", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_score, color = result), size = 3) +
  scale_color_manual(values = c("Win"="#2ecc71","Loss"="#e74c3c"), name = "Result") +
  geom_hline(yintercept = avg$pts_for, linetype="dotted", color="#2ecc71", linewidth=0.7) +
  labs(
    title    = paste(team_name_full, "— Points Scored by Week"),
    subtitle = "Green = points for  |  Red dashed = points against  |  Dotted = season average",
    x        = "Week", y = "Points"
  ) +
  theme_minimal(base_size = 13)
```

## Yards Per Game: Passing vs. Rushing

```{r}
#| label: yards-chart

yards_long <- team_data %>%
  select(week, pass_yards, rush_yards) %>%
  pivot_longer(cols = c(pass_yards, rush_yards), names_to = "Type", values_to = "Yards") %>%
  mutate(Type = recode(Type, pass_yards="Passing", rush_yards="Rushing"))

ggplot(yards_long, aes(x = week, y = Yards, color = Type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Passing"="#3498db","Rushing"="#e67e22")) +
  labs(
    title    = paste(team_name_full, "— Yards Per Game by Week"),
    x        = "Week", y = "Yards", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic = c("Points Per Game","Points Allowed Per Game","Point Differential",
                "Passing Yards Per Game","Rushing Yards Per Game","Total Yards Per Game",
                "Turnovers Per Game","Sacks Per Game (defense)","Takeaways Per Game"),
  Value     = c(
    sprintf("%.1f", avg$pts_for),
    sprintf("%.1f", avg$pts_against),
    sprintf("%+.1f", avg$pt_diff),
    sprintf("%.1f", avg$pass_yards),
    sprintf("%.1f", avg$rush_yards),
    sprintf("%.1f", avg$total_yards),
    sprintf("%.2f", avg$turnovers),
    sprintf("%.2f", avg$sacks_made),
    sprintf("%.2f", avg$def_tov)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NFL per-game averages and weighted into an overall score on a 4.0 scale. Scoring and defense each carry 18% of the grade since points are the ultimate NFL currency. Turnover margin (both offensive turnovers and defensive takeaways) accounts for a combined 22%, reflecting how heavily turnovers drive NFL outcomes.
