---
title: "MLB Team Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages

# install.packages("baseballr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr")

library(baseballr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which MLB team do you want to grade?
# Use the Baseball Reference 3-letter team abbreviation.
# Examples: "NYY", "BOS", "LAD", "HOU", "ATL", "CHC", "SFG", "NYM"
# Not sure? Run: baseballr::bref_team_results("NYY", 2024) to test a known team,
# then swap your team in.
my_team <- "NYY"

# Which season? Use the calendar year.
# Example: the 2024 season is 2024.
season_year <- 2024

# Season label shown in the report (for display only)
season_label <- "2024"

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

# Pull game-by-game results from Baseball Reference
game_log <- baseballr::bref_team_results(Tm = my_team, yr = season_year)

if (is.null(game_log) || nrow(game_log) == 0) {
  stop("No data found for '", my_team, "' in ", season_year,
       ". Check the abbreviation — Baseball Reference uses standard 3-letter codes.")
}

# Clean up the results
team_data <- game_log %>%
  filter(!is.na(R), !is.na(RA)) %>%
  mutate(
    runs_scored  = as.numeric(R),
    runs_allowed = as.numeric(RA),
    won          = grepl("^W", `W/L`),
    run_diff     = runs_scored - runs_allowed
  )

team_name_full <- paste(my_team, season_label, "Season")
games_played   <- nrow(team_data)
wins           <- sum(team_data$won, na.rm = TRUE)
losses         <- games_played - wins
win_pct        <- wins / games_played * 100

# Pull offensive team stats from FanGraphs (season totals, convert to per-game)
batting <- tryCatch(
  baseballr::fg_team_batter(startseason = as.character(season_year),
                             endseason   = as.character(season_year)) %>%
    filter(teamName == my_team | abbreviation == my_team),
  error = function(e) NULL
)

pitching <- tryCatch(
  baseballr::fg_team_pitcher(startseason = as.character(season_year),
                              endseason   = as.character(season_year)) %>%
    filter(teamName == my_team | abbreviation == my_team),
  error = function(e) NULL
)
```

```{r}
#| label: averages
#| include: false

avg_runs <- team_data %>%
  summarise(
    runs_per_game    = mean(runs_scored,  na.rm = TRUE),
    runs_allowed_pg  = mean(runs_allowed, na.rm = TRUE),
    run_diff         = mean(run_diff,     na.rm = TRUE)
  )

# FanGraphs stats (if available)
avg_bat <- if (!is.null(batting) && nrow(batting) > 0) {
  list(
    avg      = as.numeric(batting$AVG[1]),
    obp      = as.numeric(batting$OBP[1]),
    slg      = as.numeric(batting$SLG[1]),
    ops      = as.numeric(batting$OPS[1]),
    hr_pg    = as.numeric(batting$HR[1]) / games_played,
    k_pct    = as.numeric(batting$`K%`[1]) * 100,
    bb_pct   = as.numeric(batting$`BB%`[1]) * 100
  )
} else {
  list(avg=NA, obp=NA, slg=NA, ops=NA, hr_pg=NA, k_pct=NA, bb_pct=NA)
}

avg_pit <- if (!is.null(pitching) && nrow(pitching) > 0) {
  list(
    era   = as.numeric(pitching$ERA[1]),
    whip  = as.numeric(pitching$WHIP[1]),
    fip   = as.numeric(pitching$FIP[1]),
    k9    = as.numeric(pitching$`K/9`[1]),
    bb9   = as.numeric(pitching$`BB/9`[1])
  )
} else {
  list(era=NA, whip=NA, fip=NA, k9=NA, bb9=NA)
}
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (is.na(value) || is.nan(value)) return("N/A")
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  if (l == "N/A") return(NA_real_)
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# Benchmarks: MLB league averages (2024 season)
grade_runs_scored <- grade_stat(avg_runs$runs_per_game,   c(A=5.2, B=4.6, C=4.1, D=3.6))
grade_runs_allowed <- grade_stat(avg_runs$runs_allowed_pg, c(A=3.6, B=4.1, C=4.6, D=5.2), lower_is_better=TRUE)
grade_ops         <- grade_stat(avg_bat$ops,              c(A=.780, B=.740, C=.700, D=.660))
grade_obp         <- grade_stat(avg_bat$obp,              c(A=.335, B=.318, C=.305, D=.290))
grade_slg         <- grade_stat(avg_bat$slg,              c(A=.445, B=.415, C=.385, D=.360))
grade_era         <- grade_stat(avg_pit$era,              c(A=3.4, B=3.9, C=4.4, D=5.0),  lower_is_better=TRUE)
grade_whip        <- grade_stat(avg_pit$whip,             c(A=1.12, B=1.22, C=1.32, D=1.42), lower_is_better=TRUE)
grade_win_pct     <- grade_stat(win_pct,                  c(A=60, B=54, C=49, D=43))

all_grades <- c(grade_runs_scored, grade_runs_allowed, grade_ops, grade_obp,
                grade_slg, grade_era, grade_whip, grade_win_pct)
weights    <- c(0.15, 0.15, 0.14, 0.10, 0.10, 0.16, 0.12, 0.08)

valid      <- !is.na(sapply(all_grades, letter_to_gpa))
overall_gpa   <- weighted.mean(sapply(all_grades, letter_to_gpa)[valid], weights[valid])
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Runs Scored/G","Runs Allowed/G","OPS","OBP","SLG","ERA","WHIP","Win %"),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
) %>% filter(!is.na(GPA))

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r my_team` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) | **Run Differential:** `r sprintf("%+.1f", avg_runs$run_diff)` per game | **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(my_team, "— Category Grades"),
    subtitle = paste(season_label, "MLB Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Runs Scored vs. Runs Allowed Per Game

```{r}
#| label: runs-trend

trend_data <- team_data %>%
  mutate(game_num = row_number(),
         result   = ifelse(won, "Win", "Loss"))

ggplot(trend_data, aes(x = game_num)) +
  geom_line(aes(y = runs_scored),  color = "#2ecc71", linewidth = 0.7, alpha = 0.7) +
  geom_line(aes(y = runs_allowed), color = "#e74c3c", linewidth = 0.7, linetype="dashed", alpha = 0.7) +
  geom_smooth(aes(y = runs_scored),  color = "#2ecc71", se = FALSE, linewidth = 1.2) +
  geom_smooth(aes(y = runs_allowed), color = "#e74c3c", se = FALSE, linewidth = 1.2, linetype="dashed") +
  labs(
    title    = paste(my_team, "— Runs Per Game"),
    subtitle = "Green = runs scored  |  Red dashed = runs allowed  |  Smoothed trend lines",
    x        = "Game Number", y = "Runs"
  ) +
  theme_minimal(base_size = 13)
```

## Cumulative Win/Loss Record

```{r}
#| label: win-loss

trend_data <- trend_data %>%
  mutate(
    cum_wins   = cumsum(as.integer(won)),
    cum_losses = cumsum(as.integer(!won))
  )

trend_long <- trend_data %>%
  select(game_num, cum_wins, cum_losses) %>%
  pivot_longer(cols = c(cum_wins, cum_losses), names_to = "Type", values_to = "Count") %>%
  mutate(Type = recode(Type, cum_wins = "Wins", cum_losses = "Losses"))

ggplot(trend_long, aes(x = game_num, y = Count, color = Type)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Wins"="#2ecc71","Losses"="#e74c3c")) +
  labs(
    title = paste(my_team, "— Cumulative Win/Loss Record"),
    x = "Game Number", y = "Games", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic = c("Record","Win %","Runs Per Game","Runs Allowed Per Game",
                "Run Differential","Batting Average","OBP","SLG","OPS","ERA","WHIP"),
  Value     = c(
    paste0(wins, "-", losses),
    sprintf("%.1f%%", win_pct),
    sprintf("%.2f", avg_runs$runs_per_game),
    sprintf("%.2f", avg_runs$runs_allowed_pg),
    sprintf("%+.2f", avg_runs$run_diff),
    ifelse(is.na(avg_bat$avg),  "N/A", sprintf(".%03d", round(avg_bat$avg * 1000))),
    ifelse(is.na(avg_bat$obp),  "N/A", sprintf(".%03d", round(avg_bat$obp * 1000))),
    ifelse(is.na(avg_bat$slg),  "N/A", sprintf(".%03d", round(avg_bat$slg * 1000))),
    ifelse(is.na(avg_bat$ops),  "N/A", sprintf(".%03d", round(avg_bat$ops * 1000))),
    ifelse(is.na(avg_pit$era),  "N/A", sprintf("%.2f", avg_pit$era)),
    ifelse(is.na(avg_pit$whip), "N/A", sprintf("%.2f", avg_pit$whip))
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Value"))
```

## How the Grade Is Calculated

Each category is graded against MLB league averages from the 2024 season and weighted into an overall score. ERA carries the most weight (16%) because pitching is the dominant factor in baseball outcomes. Runs scored and runs allowed each carry 15%. OPS combines on-base and slugging to reflect overall offensive quality in a single number.

Game-by-game results come from Baseball Reference. Batting and pitching percentage stats (OPS, ERA, WHIP) come from FanGraphs — if those didn't load, those categories will show as N/A and only the run-based grades contribute to the overall.
