---
title: "NHL Team Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages

 install.packages("pak")
 pak::pak("sportsdataverse/hockeyR")
 install.packages("dplyr")
 install.packages("ggplot2")
 install.packages("tidyr")

library(hockeyR)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which NHL team do you want to grade?
# Use the 3-letter team abbreviation ESPN / NHL use.
# Examples: "BOS", "NYR", "TOR", "EDM", "COL", "VGK", "TBL"
# Not sure of yours? Run:  hockeyR::team_logos_colors %>% select(team_abbr, full_team_name) %>% print(n=40)
my_team <- "BOS"

# Which season? Use the year the season STARTS.
# Example: the 2024-25 season is written as 2024.
season_year <- 2024

# Season label shown in the report (for display only)
season_label <- "2024-25"

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

# Load play-by-play and aggregate to game-level team stats
pbp <- hockeyR::load_pbp(seasons = season_year)

# Get final scores per game
game_scores <- pbp %>%
  group_by(game_id, home_abbreviation, away_abbreviation) %>%
  summarise(
    home_goals = max(home_final, na.rm = TRUE),
    away_goals = max(away_final, na.rm = TRUE),
    home_shots = sum(event_type %in% c("SHOT","GOAL") & team == home_abbreviation, na.rm = TRUE),
    away_shots = sum(event_type %in% c("SHOT","GOAL") & team == away_abbreviation, na.rm = TRUE),
    home_pp_goals = sum(event_type == "GOAL" & team == home_abbreviation &
                          strength_code %in% c("PP","PP2"), na.rm = TRUE),
    away_pp_goals = sum(event_type == "GOAL" & team == away_abbreviation &
                          strength_code %in% c("PP","PP2"), na.rm = TRUE),
    home_pp_opps = sum(event_type == "PENALTY" & team == away_abbreviation, na.rm = TRUE),
    away_pp_opps = sum(event_type == "PENALTY" & team == home_abbreviation, na.rm = TRUE),
    .groups = "drop"
  )

# Home games
home_games <- game_scores %>%
  filter(home_abbreviation == my_team) %>%
  transmute(
    game_id,
    team_goals   = home_goals,
    opp_goals    = away_goals,
    team_shots   = home_shots,
    opp_shots    = away_shots,
    pp_goals     = home_pp_goals,
    pp_opps      = home_pp_opps,
    pk_goals_against = away_pp_goals,
    pk_opps      = away_pp_opps,
    won          = home_goals > away_goals
  )

# Away games
away_games <- game_scores %>%
  filter(away_abbreviation == my_team) %>%
  transmute(
    game_id,
    team_goals   = away_goals,
    opp_goals    = home_goals,
    team_shots   = away_shots,
    opp_shots    = home_shots,
    pp_goals     = away_pp_goals,
    pp_opps      = away_pp_opps,
    pk_goals_against = home_pp_goals,
    pk_opps      = home_pp_opps,
    won          = away_goals > home_goals
  )

team_data <- bind_rows(home_games, away_games)

if (nrow(team_data) == 0) stop("No games found for '", my_team, "'. Check the abbreviation.")

# Look up full team name
team_name_full <- tryCatch(
  hockeyR::team_logos_colors %>%
    filter(team_abbr == my_team) %>%
    pull(full_team_name) %>% first(),
  error = function(e) my_team
)

games_played <- nrow(team_data)
wins         <- sum(team_data$won, na.rm = TRUE)
losses       <- games_played - wins
win_pct      <- wins / games_played * 100
```

```{r}
#| label: averages
#| include: false

avg <- team_data %>%
  mutate(
    pp_pct = ifelse(pp_opps > 0, pp_goals / pp_opps * 100, NA),
    pk_pct = ifelse(pk_opps > 0, (1 - pk_goals_against / pk_opps) * 100, NA),
    shot_diff = team_shots - opp_shots
  ) %>%
  summarise(
    goals_for    = mean(team_goals,  na.rm = TRUE),
    goals_against = mean(opp_goals,  na.rm = TRUE),
    shots_for    = mean(team_shots,  na.rm = TRUE),
    shots_against = mean(opp_shots,  na.rm = TRUE),
    pp_pct       = mean(pp_pct,      na.rm = TRUE),
    pk_pct       = mean(pk_pct,      na.rm = TRUE),
    shot_diff    = mean(shot_diff,   na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# Benchmarks: NHL league averages (2024-25 season)
grade_offense   <- grade_stat(avg$goals_for,    c(A=3.4, B=3.1, C=2.8, D=2.5))
grade_defense   <- grade_stat(avg$goals_against, c(A=2.5, B=2.8, C=3.1, D=3.4), lower_is_better=TRUE)
grade_shots     <- grade_stat(avg$shots_for,    c(A=34,  B=31,  C=28,  D=25))
grade_shot_supp <- grade_stat(avg$shots_against, c(A=25,  B=28,  C=31,  D=34),  lower_is_better=TRUE)
grade_pp        <- grade_stat(avg$pp_pct,        c(A=24,  B=21,  C=18,  D=15))
grade_pk        <- grade_stat(avg$pk_pct,        c(A=85,  B=82,  C=79,  D=76))
grade_win_pct   <- grade_stat(win_pct,           c(A=62,  B=55,  C=48,  D=40))

all_grades <- c(grade_offense, grade_defense, grade_shots,
                grade_shot_supp, grade_pp, grade_pk, grade_win_pct)
weights    <- c(0.20, 0.20, 0.12, 0.12, 0.14, 0.12, 0.10)

overall_gpa   <- sum(sapply(all_grades, letter_to_gpa) * weights)
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Offense (GF/G)","Defense (GA/G)","Shots For","Shots Against",
               "Power Play %","Penalty Kill %","Win %"),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
)

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r team_name_full` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) \| **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

------------------------------------------------------------------------

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(team_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NHL Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Goals Scored vs. Goals Allowed Per Game

```{r}
#| label: goals-trend

trend_data <- team_data %>%
  mutate(game_num = row_number(),
         result   = ifelse(won, "Win", "Loss"))

ggplot(trend_data, aes(x = game_num)) +
  geom_line(aes(y = team_goals), color = "#2ecc71", linewidth = 1) +
  geom_line(aes(y = opp_goals),  color = "#e74c3c", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_goals, color = result), size = 2.5) +
  scale_color_manual(values = c("Win"="#2ecc71","Loss"="#e74c3c"), name = "Result") +
  geom_hline(yintercept = avg$goals_for,    linetype="dotted", color="#2ecc71", linewidth=0.7) +
  geom_hline(yintercept = avg$goals_against, linetype="dotted", color="#e74c3c", linewidth=0.7) +
  labs(
    title    = paste(team_name_full, "— Goals Per Game"),
    subtitle = "Green = goals for  |  Red dashed = goals against  |  Dotted = season averages",
    x        = "Game Number", y = "Goals"
  ) +
  theme_minimal(base_size = 13)
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic = c("Goals Per Game","Goals Against Per Game","Shots Per Game",
                "Shots Against Per Game","Power Play %","Penalty Kill %",
                "Shot Differential","Win %"),
  Value     = c(
    sprintf("%.2f", avg$goals_for),
    sprintf("%.2f", avg$goals_against),
    sprintf("%.1f", avg$shots_for),
    sprintf("%.1f", avg$shots_against),
    sprintf("%.1f%%", avg$pp_pct),
    sprintf("%.1f%%", avg$pk_pct),
    sprintf("%+.1f", avg$shot_diff),
    sprintf("%.1f%%", win_pct)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NHL league averages and weighted into an overall score on a 4.0 scale. Offense and defense (goals for/against) each carry 20% of the overall grade. Power play percentage carries 14% because special teams are a major driver of NHL success.
