---
title: "NCAA Player Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages
library(hoopR)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which player do you want to grade?
# Use the player's full name exactly as ESPN lists it (First Last).
# Examples: "Cooper Flagg", "Zach Edey", "Hunter Dickinson"
my_player <- "Cooper Flagg"

# Which season? Use the year the season ENDS.
# Example: the 2024-25 season is written as 2025.
season_year <- 2025

# Season label shown in the report (for display only)
season_label <- "2024-25"

# Player position — used to set rebounding and assist benchmarks.
# Choose ONE of: "Guard"  |  "Wing/Forward"  |  "Big/Center"
player_position <- "Wing/Forward"

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

player_box <- hoopR::load_mbb_player_box(seasons = season_year)

player_data <- player_box %>%
  filter(athlete_display_name == my_player,
         did_not_play == FALSE | is.na(did_not_play))

if (nrow(player_data) == 0) {
  stop("No games found for '", my_player, "'. Check spelling or run: ",
       "player_box %>% filter(grepl(my_player, athlete_display_name, ignore.case=TRUE)) ",
       "%>% distinct(athlete_display_name, team_location) %>% print(n=50)")
}

player_name_full <- player_data$athlete_display_name[1]
team_name_full   <- player_data$team_display_name[1]
games_played     <- nrow(player_data)
```

```{r}
#| label: averages
#| include: false

player_data <- player_data %>%
  mutate(
    pts  = as.numeric(points),
    reb  = as.numeric(rebounds),
    ast  = as.numeric(assists),
    stl  = as.numeric(steals),
    blk  = as.numeric(blocks),
    tov  = as.numeric(turnovers),
    fgm  = as.numeric(field_goals_made),
    fga  = as.numeric(field_goals_attempted),
    ftm  = as.numeric(free_throws_made),
    fta  = as.numeric(free_throws_attempted),
    tpm  = as.numeric(three_point_field_goals_made),
    tpa  = as.numeric(three_point_field_goals_attempted),
    min  = as.numeric(minutes),
    game_date = as.Date(game_date)
  ) %>%
  mutate(
    fg_pct    = ifelse(fga > 0, fgm / fga * 100, NA),
    three_pct = ifelse(tpa > 0, tpm / tpa * 100, NA),
    ft_pct    = ifelse(fta > 0, ftm / fta * 100, NA),
    efficiency = pts + reb + ast + stl + blk - tov - (fga - fgm)
  )

avg <- player_data %>%
  summarise(
    pts       = mean(pts,       na.rm = TRUE),
    reb       = mean(reb,       na.rm = TRUE),
    ast       = mean(ast,       na.rm = TRUE),
    stl       = mean(stl,       na.rm = TRUE),
    blk       = mean(blk,       na.rm = TRUE),
    tov       = mean(tov,       na.rm = TRUE),
    fg_pct    = mean(fg_pct,    na.rm = TRUE),
    three_pct = mean(three_pct, na.rm = TRUE),
    ft_pct    = mean(ft_pct,    na.rm = TRUE),
    min       = mean(min,       na.rm = TRUE),
    efficiency = mean(efficiency, na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

reb_thresholds <- switch(player_position,
  "Guard"        = c(A=6,  B=4,  C=3,  D=2),
  "Wing/Forward" = c(A=8,  B=6,  C=4,  D=2),
  "Big/Center"   = c(A=10, B=8,  C=6,  D=4)
)
ast_thresholds <- switch(player_position,
  "Guard"        = c(A=6,  B=4,   C=2.5, D=1.5),
  "Wing/Forward" = c(A=4,  B=2.5, C=1.5, D=0.8),
  "Big/Center"   = c(A=2.5,B=1.5, C=0.8, D=0.3)
)

grade_scoring    <- grade_stat(avg$pts,              c(A=20, B=16, C=12, D=8))
grade_rebounding <- grade_stat(avg$reb,              reb_thresholds)
grade_assists    <- grade_stat(avg$ast,              ast_thresholds)
grade_defense    <- grade_stat(avg$stl + avg$blk,   c(A=3.0,B=2.0,C=1.2,D=0.6))
grade_turnovers  <- grade_stat(avg$tov,              c(A=1.5,B=2.5,C=3.5,D=4.5), lower_is_better=TRUE)
grade_fg         <- grade_stat(avg$fg_pct,           c(A=50, B=45, C=40, D=35))
grade_ft         <- grade_stat(avg$ft_pct,           c(A=80, B=73, C=65, D=55))
grade_efficiency <- grade_stat(avg$efficiency,       c(A=20, B=15, C=10, D=5))

all_grades <- c(grade_scoring, grade_rebounding, grade_assists, grade_defense,
                grade_turnovers, grade_fg, grade_ft, grade_efficiency)
weights    <- c(0.22, 0.15, 0.13, 0.12, 0.10, 0.12, 0.08, 0.08)

overall_gpa   <- sum(sapply(all_grades, letter_to_gpa) * weights)
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Scoring","Rebounding","Assists","Defense (stl+blk)",
               "Ball Security","Field Goal %","Free Throw %","Efficiency"),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
)

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r player_name_full` — `r team_name_full` | `r season_label`

**Position:** `r player_position` | **Games Graded:** `r games_played` | **Minutes Per Game:** `r sprintf("%.1f", avg$min)`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(player_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NCAA Season |", player_position),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Scoring Trend

```{r}
#| label: scoring-trend

trend_data <- player_data %>%
  arrange(game_date) %>%
  mutate(game_num = row_number())

ggplot(trend_data, aes(x = game_num, y = pts)) +
  geom_line(color = "#3498db", linewidth = 1) +
  geom_point(color = "#3498db", size = 2.5) +
  geom_hline(yintercept = avg$pts, linetype = "dashed", color = "#e74c3c", linewidth = 0.8) +
  annotate("text", x = max(trend_data$game_num), y = avg$pts + 0.8,
           label = paste("Avg:", sprintf("%.1f", avg$pts)), color = "#e74c3c", hjust = 1, size = 3.5) +
  labs(
    title    = paste(player_name_full, "— Points Per Game"),
    subtitle = "Red dashed line = season average",
    x        = "Game Number",
    y        = "Points"
  ) +
  theme_minimal(base_size = 13)
```

## Game-by-Game Stats: Points, Rebounds, Assists

```{r}
#| label: pra-chart

pra_long <- trend_data %>%
  select(game_num, pts, reb, ast) %>%
  pivot_longer(cols = c(pts, reb, ast), names_to = "Stat", values_to = "Value") %>%
  mutate(Stat = recode(Stat, pts="Points", reb="Rebounds", ast="Assists"))

ggplot(pra_long, aes(x = game_num, y = Value, color = Stat)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Points"="#3498db","Rebounds"="#2ecc71","Assists"="#e67e22")) +
  labs(
    title    = paste(player_name_full, "— Points / Rebounds / Assists by Game"),
    x        = "Game Number",
    y        = "Count",
    color    = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic = c("Points Per Game","Rebounds Per Game","Assists Per Game",
                "Steals Per Game","Blocks Per Game","Turnovers Per Game",
                "Field Goal %","Three-Point %","Free Throw %","Efficiency Index"),
  Value     = c(
    sprintf("%.1f", avg$pts),
    sprintf("%.1f", avg$reb),
    sprintf("%.1f", avg$ast),
    sprintf("%.1f", avg$stl),
    sprintf("%.1f", avg$blk),
    sprintf("%.1f", avg$tov),
    sprintf("%.1f%%", avg$fg_pct),
    sprintf("%.1f%%", avg$three_pct),
    sprintf("%.1f%%", avg$ft_pct),
    sprintf("%.1f",  avg$efficiency)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NCAA Division I starter averages and converted to a 4.0 scale. Scoring carries the most weight (22%), followed by field goal percentage (12%), rebounding (15%), and assists (13%). Rebounding and assist benchmarks adjust automatically based on the position you set at the top of the file.

The Efficiency Index adds together points, rebounds, assists, steals, and blocks, then subtracts turnovers and missed field goals — a simple measure of overall contribution per game.
