---
title: "NBA Team Season Grade Report"
author: "Pat Rankin"
format: html
editor: visual
execute:
  message: false
  warning: false
---

```{r}
#| label: packages
library(hoopR)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```

## ⚙️ Your Settings — Change These Before Rendering

```{r}
#| label: settings

# ============================================================
#  FILL IN YOUR INFO BELOW — this is the only section
#  you need to change each time you run this report.
# ============================================================

# Which NBA team do you want to grade?
# Use the team's common name exactly as ESPN lists it.
# Examples: "Lakers", "Celtics", "Warriors", "Knicks", "Heat"
my_team <- "Lakers"

# Which season? Use the year the season ENDS.
# Example: the 2025-26 NBA season is written as 2026.
season_year <- 2026

# Season label shown in the report (for display only)
season_label <- "2025-26"

# ============================================================
#  THAT'S IT — scroll down to render and read your report.
# ============================================================
```

```{r}
#| label: ingest
#| include: false

team_box <- hoopR::load_nba_team_box(seasons = season_year)

team_data <- team_box %>%
  filter(team_location == my_team |
           team_name == my_team |
           team_short_display_name == my_team)

if (nrow(team_data) == 0) {
  stop("No games found for '", my_team, "'. Check spelling or run: ",
       "team_box %>% distinct(team_location) %>% arrange(team_location) %>% print(n=40)")
}

team_name_full <- team_data$team_display_name[1]
games_played   <- nrow(team_data)
wins           <- sum(team_data$team_winner == TRUE,  na.rm = TRUE)
losses         <- sum(team_data$team_winner == FALSE, na.rm = TRUE)
win_pct        <- wins / games_played * 100
```

```{r}
#| label: averages
#| include: false

avg <- team_data %>%
  mutate(
    team_score = as.numeric(team_score),
    opp_score  = as.numeric(opponent_team_score),
    fg_pct     = as.numeric(field_goal_pct)             * 100,
    three_pct  = as.numeric(three_point_field_goal_pct) * 100,
    ft_pct     = as.numeric(free_throw_pct)             * 100,
    total_reb  = as.numeric(total_rebounds),
    ast        = as.numeric(assists),
    stl        = as.numeric(steals),
    blk        = as.numeric(blocks),
    tov        = as.numeric(turnovers),
    pt_diff    = team_score - opp_score
  ) %>%
  summarise(
    pts       = mean(team_score, na.rm = TRUE),
    opp_pts   = mean(opp_score,  na.rm = TRUE),
    fg_pct    = mean(fg_pct,     na.rm = TRUE),
    three_pct = mean(three_pct,  na.rm = TRUE),
    ft_pct    = mean(ft_pct,     na.rm = TRUE),
    reb       = mean(total_reb,  na.rm = TRUE),
    ast       = mean(ast,        na.rm = TRUE),
    stl       = mean(stl,        na.rm = TRUE),
    blk       = mean(blk,        na.rm = TRUE),
    tov       = mean(tov,        na.rm = TRUE),
    pt_diff   = mean(pt_diff,    na.rm = TRUE)
  )
```

```{r}
#| label: grading
#| include: false

score_to_letter <- function(score) {
  if      (score >= 4.0) "A+"
  else if (score >= 3.7) "A"
  else if (score >= 3.3) "A-"
  else if (score >= 3.0) "B+"
  else if (score >= 2.7) "B"
  else if (score >= 2.3) "B-"
  else if (score >= 2.0) "C+"
  else if (score >= 1.7) "C"
  else if (score >= 1.3) "C-"
  else if (score >= 1.0) "D+"
  else if (score >= 0.7) "D"
  else if (score >= 0.3) "D-"
  else                   "F"
}

grade_stat <- function(value, thresholds, lower_is_better = FALSE) {
  if (lower_is_better) {
    if      (value <= thresholds["A"]) gpa <- 4.0
    else if (value <= thresholds["B"]) gpa <- 3.0
    else if (value <= thresholds["C"]) gpa <- 2.0
    else if (value <= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  } else {
    if      (value >= thresholds["A"]) gpa <- 4.0
    else if (value >= thresholds["B"]) gpa <- 3.0
    else if (value >= thresholds["C"]) gpa <- 2.0
    else if (value >= thresholds["D"]) gpa <- 1.0
    else                               gpa <- 0.0
  }
  score_to_letter(gpa)
}

letter_to_gpa <- function(l) {
  switch(l,
    "A+"=4.3,"A"=4.0,"A-"=3.7,
    "B+"=3.3,"B"=3.0,"B-"=2.7,
    "C+"=2.3,"C"=2.0,"C-"=1.7,
    "D+"=1.3,"D"=1.0,"D-"=0.7,"F"=0.0)
}

# Benchmarks: NBA league averages (2024-25 season)
grade_scoring    <- grade_stat(avg$pts,       c(A=118,B=113,C=108,D=103))
grade_defense    <- grade_stat(avg$opp_pts,   c(A=103,B=108,C=113,D=118), lower_is_better=TRUE)
grade_fg         <- grade_stat(avg$fg_pct,    c(A=50, B=47, C=44, D=41))
grade_three      <- grade_stat(avg$three_pct, c(A=38, B=36, C=34, D=32))
grade_ft         <- grade_stat(avg$ft_pct,    c(A=80, B=76, C=72, D=68))
grade_rebounding <- grade_stat(avg$reb,       c(A=46, B=43, C=41, D=38))
grade_assists    <- grade_stat(avg$ast,       c(A=28, B=25, C=22, D=19))
grade_turnovers  <- grade_stat(avg$tov,       c(A=12, B=13.5,C=15,D=16.5), lower_is_better=TRUE)
grade_win_pct    <- grade_stat(win_pct,       c(A=65, B=55, C=45, D=35))

all_grades <- c(grade_scoring, grade_defense, grade_fg, grade_three,
                grade_ft, grade_rebounding, grade_assists,
                grade_turnovers, grade_win_pct)
weights    <- c(0.15, 0.15, 0.12, 0.10, 0.08, 0.10, 0.10, 0.10, 0.10)

overall_gpa   <- sum(sapply(all_grades, letter_to_gpa) * weights)
overall_grade <- score_to_letter(overall_gpa)

grade_df <- data.frame(
  Category = c("Scoring","Defense","FG%","3-Point%",
               "Free Throw%","Rebounding","Assists",
               "Ball Security","Win %"),
  Grade    = all_grades,
  GPA      = sapply(all_grades, letter_to_gpa)
)

grade_df$Grade_Group <- case_when(
  grade_df$GPA >= 3.7 ~ "A",
  grade_df$GPA >= 2.7 ~ "B",
  grade_df$GPA >= 1.7 ~ "C",
  grade_df$GPA >= 0.7 ~ "D",
  TRUE                ~ "F"
)
```

## `r team_name_full` — `r season_label` Season Report

**Record:** `r wins`-`r losses` (`r sprintf("%.1f", win_pct)`% win rate) | **Point Differential:** `r sprintf("%+.1f", avg$pt_diff)` per game | **Games Graded:** `r games_played`

**Overall Season Grade: `r overall_grade`**

---

## Category Grades

```{r}
#| label: grade-chart

grade_colors <- c("A"="#2ecc71","B"="#3498db","C"="#f39c12","D"="#e67e22","F"="#e74c3c")

ggplot(grade_df, aes(x = reorder(Category, GPA), y = GPA, fill = Grade_Group)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Grade), hjust = -0.2, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = grade_colors, name = "Grade") +
  scale_y_continuous(limits = c(0, 5.2), breaks = c(0,1,2,3,4)) +
  coord_flip() +
  labs(
    title    = paste(team_name_full, "— Category Grades"),
    subtitle = paste(season_label, "NBA Season"),
    x        = NULL,
    y        = "Grade Points (4.0 scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

## Scoring Trend

```{r}
#| label: scoring-trend

trend_data <- team_data %>%
  mutate(
    game_date  = as.Date(game_date),
    team_score = as.numeric(team_score),
    opp_score  = as.numeric(opponent_team_score),
    result     = ifelse(team_winner == TRUE, "Win", "Loss")
  ) %>%
  arrange(game_date) %>%
  mutate(game_num = row_number())

ggplot(trend_data, aes(x = game_num)) +
  geom_line(aes(y = team_score), color = "#2ecc71", linewidth = 1) +
  geom_line(aes(y = opp_score),  color = "#e74c3c", linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = team_score, color = result), size = 2) +
  scale_color_manual(values = c("Win"="#2ecc71","Loss"="#e74c3c"), name = "Result") +
  geom_hline(yintercept = avg$pts, linetype = "dotted", color = "#2ecc71", linewidth = 0.7) +
  labs(
    title    = paste(team_name_full, "— Scoring by Game"),
    subtitle = "Green = team score  |  Red dashed = opponent score  |  Dotted = season average",
    x        = "Game Number",
    y        = "Points"
  ) +
  theme_minimal(base_size = 13)
```

## Win / Loss Rolling Record

```{r}
#| label: win-loss

trend_data <- trend_data %>%
  mutate(win_flag = ifelse(result == "Win", 1, 0),
         cumulative_wins   = cumsum(win_flag),
         cumulative_losses = cumsum(1 - win_flag))

trend_long <- trend_data %>%
  select(game_num, cumulative_wins, cumulative_losses) %>%
  pivot_longer(cols = c(cumulative_wins, cumulative_losses),
               names_to = "Type", values_to = "Count") %>%
  mutate(Type = recode(Type, cumulative_wins="Wins", cumulative_losses="Losses"))

ggplot(trend_long, aes(x = game_num, y = Count, color = Type)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Wins"="#2ecc71","Losses"="#e74c3c")) +
  labs(
    title    = paste(team_name_full, "— Cumulative Win/Loss Record"),
    x        = "Game Number",
    y        = "Games",
    color    = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Key Stats Summary

```{r}
#| label: stats-table

summary_tbl <- data.frame(
  Statistic = c("Points Per Game","Opp. Points Per Game","Point Differential",
                "Field Goal %","Three-Point %","Free Throw %",
                "Rebounds Per Game","Assists Per Game",
                "Steals Per Game","Blocks Per Game","Turnovers Per Game"),
  Value     = c(
    sprintf("%.1f",   avg$pts),
    sprintf("%.1f",   avg$opp_pts),
    sprintf("%+.1f",  avg$pt_diff),
    sprintf("%.1f%%", avg$fg_pct),
    sprintf("%.1f%%", avg$three_pct),
    sprintf("%.1f%%", avg$ft_pct),
    sprintf("%.1f",   avg$reb),
    sprintf("%.1f",   avg$ast),
    sprintf("%.1f",   avg$stl),
    sprintf("%.1f",   avg$blk),
    sprintf("%.1f",   avg$tov)
  )
)

knitr::kable(summary_tbl, align = c("l","r"), col.names = c("Statistic","Season Avg"))
```

## How the Grade Is Calculated

Each category is graded against NBA league averages from the 2024-25 season and converted to a 4.0 scale. Scoring and defense each carry 15% of the overall grade. Field goal percentage accounts for 12%, while win percentage, ball security, rebounding, and assists each account for 10%.

The grade scale is: **A+** (elite) · **A / A-** (excellent) · **B** range (above average) · **C** range (league average) · **D** range (below average) · **F** (well below average).
